---
output:
  html_document: 
    theme: cerulean
    toc: true
    toc_float: true
    toc_depth: 4
---




```{r setKnitrOptions, include=FALSE}
knitr::opts_chunk$set( results='asis', echo = FALSE, error = TRUE, cache = FALSE, cache.comments = TRUE, fig.dim = c(7,7), autodep = TRUE)
```




```{r loadParamsAndDependencies, warning=FALSE, message=FALSE}

### Load general Deps
cat("<BR>WORKING_DIR=", WORKING_DIR)
source( file.path( WORKING_DIR, "00_generalDeps.R"));

# Import chunks from source files
read_chunk( path=file.path( WORKING_DIR, "01_prepareData.R"));
source_python( file.path( WORKING_DIR, "02_executescAANet.py"));
read_chunk( path=file.path( WORKING_DIR, "03_studyArchetypeUsage.R"));
read_chunk( path=file.path( WORKING_DIR, "04a_studyArchetypeEnrichment_GLM.R"));
read_chunk( path=file.path( WORKING_DIR, "04b_studyArchetypeEnrichment_seurat.R"));
read_chunk( path=file.path( WORKING_DIR, "04c_studyArchetypeEnrichment_scShape.R"));
read_chunk( path=file.path( WORKING_DIR, "90_finalize.R"));
```

---
title: `r paste( SCIENTIFIC_PROJECT_NAME, ANALYSIS_STEP_LITTERAL_DESCRIPTION, sep =" - ")`
---


```{r initFunctions}

### Create output directory
dir.create( PATH_ANALYSIS_OUTPUT, recursive = TRUE, showWarnings = FALSE);


### Record starting time (for report and file names)
startTime = Sys.time();
startTimeReport   = format( startTime, "%Y %m %d %H:%M:%S")
startTimeFileName = format( startTime, "%Y_%m_%d_%H_%M_%S")


### Initialize utils

# Create a hash for all parameters
paramsHash = digest( paramsEnv, algo="murmur32");

# Create a 'safe' unique prefix for output files
outputFilesPrefix = path_sanitize( gsub("([[:punct:]])|\\s+", 
                                        "_", 
                                        paste0( SCIENTIFIC_PROJECT_NAME, "_",
                                                EXPERIMENT_NAME, "_")),
                                   replacement = "_");

# Set seed for pseudo random number generator
set.seed(42);

# Create a hidden empty heatmap to prevent heatmapr interference with plotly functions (bug ?)
cat('<div style="display: none">');
# main_heatmap(matrix(0));
cat('</div>');
```




## Prepare data for DEG analysis


### Run informations {.tabset .tabset-fade}

#### Run
<center>
```{r runInfo_run}
pander( c( 'Date'     = startTimeReport,
           'Hash'     = paramsHash));
```
</center>

#### Parameters
```{r runInfo_params}
listParams = as.list( paramsEnv);
pander(listParams[order(names(listParams))]); # Show params in alphabetical order
```

#### System
<center>
```{r runInfo_sysInfo}
as.data.frame( Sys.info()) %>% kbl() %>% kable_styling()
```
</center>

#### R
```{r runInfo_r}
pander( sessionInfo());
```

<!-- Ending the tabset section -->
### {.toc-ignore} 
<br /><br />




### Load data from previous analysis step and prepare it for scAANet
```{r prepareData}
<<prepare_data>>
```
<br /><br />


### Execute scAANet on data
```{python scAANetAnalysis}

import os
import pandas as pd
import numpy as np
import anndata
from scAAnet.api import scAAnet
import statsmodels.formula.api
# import statsmodels.api

# Get the transcriptomi expression matrix from R session
counts = pd.DataFrame( r.sc10x_counts_df)

# Define the number of archetype to find
K = int( r.AUTOENCODER_LAYER_INTERNAL_NODES)

# Define the number of nodes in the initial and final layer of the autoencoder
N = int( r.AUTOENCODER_LAYER_IN_OUT_NODES)

# Define the path to the file of the archetype analysis results
usage_file = os.path.join( r.PATH_ANALYSIS_OUTPUT, "archetype_usage_K" + str( K) + "_N" + str( N) + ".csv")
spectra_file = os.path.join( r.PATH_ANALYSIS_OUTPUT, "archetype_spectra_K" + str( K) + "_N" + str( N) + ".csv")
recon_file =  os.path.join( r.PATH_ANALYSIS_OUTPUT, "archetype_recon_K" + str( K) + "_N" + str( N) + ".csv")
genescore_file = os.path.join( r.PATH_ANALYSIS_OUTPUT, "archetype_genescore_K" + str( K) + "_N" + str( N) + ".csv")
deg_file = os.path.join( r.PATH_ANALYSIS_OUTPUT, "archetype_deg_K" + str( K) + "_N" + str( N) + ".csv")

# If the archetype analysis results files are present, use them
if( os.path.exists( usage_file) and os.path.exists( usage_file) and os.path.exists( usage_file)):
  
  usage = pd.read_csv( usage_file, index_col=0).apply( pd.to_numeric)
  spectra = pd.read_csv( spectra_file, index_col=0).apply( pd.to_numeric)
  if( K != len( usage.columns)):
    print( "ERROR: Loaded usage file does not contain the right number of dimension")

else:

  # Compute the latent space representation
  preds = scAAnet( counts, hidden_size=(N, K, N), ae_type='nb', epochs=200, batch_size=64, early_stop=100, reduce_lr=10, learning_rate=0.01)
  
  # Retrieve the result matrices
  recon, usage, spectra = preds['recon'], preds['usage'], preds['spectra']
  
  # Assign the index names to matrices
  usage = pd.DataFrame( usage, index=counts.index, columns = np.arange(1,K+1))
  spectra = pd.DataFrame( np.transpose( spectra), index=counts.columns, columns=np.arange(1,K+1))
  recon = pd.DataFrame( recon, index=counts.index, columns=counts.columns)
  
  # Write result matrix to CSV files
  usage.to_csv( usage_file)
  spectra.to_csv( spectra_file)
  recon.to_csv( recon_file)

# Calculate normalized gene scores, scale them and export result to file
gene_score = spectra*(np.log(spectra+1e-5).sub(np.mean(np.log(spectra+1e-5), axis=1), axis='index'))
gene_score = gene_score.sub(gene_score.min(axis=0), axis='columns').div(gene_score.max(axis=0)-gene_score.min(axis=0), axis='columns')
gene_score.to_csv( genescore_file)

# Run DEG tests to identify program-specific DEGs
# deg_results = programDEG( counts, usage.to_numpy(), test_use='nb_naive', offset=True)
# deg_results.to_csv( deg_file)

```
<br /><br />

### Study archetype usage {.tabset .tabset-fade}
```{r studyArchetypeUsage, warning=FALSE}
<<study_archetype_usage>>
```
<br /><br />

### Study archetype gene module enrichment (with Seurat method) {.tabset .tabset-fade}
```{r studyArchetypeEncrichmentSeurat, warning=FALSE}
<<study_archetype_enrichment_seurat>>
```
<br /><br />

### Study archetype gene module enrichment (with scShape method) {.tabset .tabset-fade}
```{r study_archetypeEnrichmentSCshape, warning=FALSE}
#<<study_archetype_enrichment_scshape>>
```

### Study archetype gene module enrichment (with GLM method) {.tabset .tabset-fade}
```{r studyArchetypeEncrichmentGLM, warning=FALSE}
<<study_archetype_enrichment_glm>>
```
<br /><br />

```{r saveFinalSeuratObject, warning=FALSE}
<<save_final_seurat_object>>
```

